What is an Ancestral Recombination Graph?

### 2022-04-06 paper.tex
Change paper description of fig 5
And highlight one of the trees

Given a GARG $E$ and the set of samples $S$ we can \emph{resolve} its
ancestry by noting that the resolved interval $Ir$ on an edge $(c,p,I)$
is given by [Rough notes]

\begin{itemize}
\item The ``ancestry interval'' $A_u$ of a node $u$ is the union of
all $R_{u, v}$ for all the children $v$ of $u$. (this isn't quite right)
\item The resolved interval is then the intersection of $A_u$ and the
original inheritance interval $I_v$ on the edge.
\item It would be good of if differentiate between the inheritance intervals
 from the GARG and the ancestry intervals of the resolved GARG.
\end{itemize}

@@

Not all nodes in an ARG are equal. Assuming that we have sequenced the
genomes of the samples, there are intrinsic limits on what can
be known about more ancient nodes, determined by the graph topology
and the flow ancestral material through it.
\emph{Simplification}~\citep{kelleher2018efficient} is the process
of removing nodes and re-writing edges in a GARG to remove
various types of redundancy, while maintaining the crucial properties
that the meaningful topology in the local trees is not affected,
and the relationship between nodes in adjacent trees is maintained.

\begin{figure}
\centering
\vspace{5em}
\includegraphics[width=\linewidth]{illustrations/simplification}
\caption{\label{fig-simplification}
ARG simplification. In all cases the graph visualization is shown on the left
and the equivalent marginal trees on the right.
(A) An ARG simulated from a diploid Wright-Fisher
model. Note that the genome \textsf{K} has three children and the genomes
\textsf{E} and \textsf{G} simultaneously have multiple parents and multiple children;
neither could not be directly represented in a Griffiths-like EARG encoding
(see section XXX).
(B) Simplified to remove all
% trying out this terminology - can define in the text
1-connected graph components (e.g., diamonds such as \textsf{JLMN}).
(C) Remove nodes that are unary in all local trees. This removes both recombinant nodes
that have only one child (e.g. \textsf{N}) and nodes that have more
than one child (i.e. are ``common ancestors'') but never represent coalescences
in the local trees (e.g. \textsf{R}).
(D) Rewrite edges to bypass any unary nodes in local trees. Note that this loses lineage
information such as the fact that \textsf{K} is connected to \textsf{A} via \textsf{E}; it
can also create many extra roots.
}
\end{figure}

% Q: would it help to talk about ``tree vertice'' here or would it
% just be more confusing? I worry that people would think that
% they are two different things then, and it's really important
% that they get that a node-is-a-node
The ideas of ``meaningful'' local tree topologies revolve around the
presence of ``unary'' nodes: a node in the tree that has exactly
one child. For example, Fig.~\ref{fig-simplification}E shows the
local trees along the genome for a simulated ARG
(Fig.~\ref{fig-simplification}A). Consider the right-most tree,
and the path above node $D$. Before $D$ meets its common ancestor
over the interval $(93, 100]$ with the other samples it passes through
nodes $H, K, N, R$ and $S$, reflecting the corresponding
traversal through the ARG (see section XXX). Such unary nodes
in a local tree are not usually considered significant,
because they provide very little information and are not
detectable by information from the tree. % weak
The nodes in a tree usually mark the points where samples find a
\emph{common} ancestor, rather than focusing on specific
ancestors. For most purposes, we would consider an edge
directly joining sample $D$ to the node $T$ to be more
useful and informative. Simplification is the process of removing
these uninformative and (for most computational purposes)
inefficient unary paths.

The first type of simplification that we can perform is to remove
graph topology that is invisible to the samples. The most well
known example of such topology is a so-called
``diamond''~\citep{rasmussen2014genome}
in which the two parent nodes of a recombination immediately
join again into a common ancestor (e.g. $K,N,O$ and $R$ in
Fig.~\ref{fig-simplification}A).
Unless we are specifically
interested in the recombination event or ancestral genomes,
there is no information in this topology and the diamond can b
e replaced by a single edge. More generally, any
subgraph which is singly-connected in both the leafward and
rootward direction (a ``super-diamond'') is non-identifiable and can be
replaced by a single edge. This definition includes the case
of a single node that has one inbound and one outbound edge.
Fig.~\ref{fig-simplification}B shows the result of this type of
graph topology simplification.

Simplifying away diamonds will remove many unary nodes from the
local trees, there can still be nodes that are unary in all
of the local trees. For example, node $H$ is one of the
parents of the recombinant node $D$ is still present in
Fig.~\ref{fig-simplification}B after diamond removal, but
it is unary in all of the local trees since it has an
in-degree of $1$.
Fig.~\ref{fig-simplification}C shows the resulting graph
after removing these nodes.

The remaining nodes are MRCAs of some subset of the samples
at \emph{some} positions along the genome. We still have
some unary nodes in the local trees, but these nodes will
correspond to a coalescence in at least one other
local tree. For example, node $E$ is unary in the first tree
of Fig.~\ref{fig-simplification}E, but is either binary
or ternary in the other trees (recall this is a Wright-Fisher
simulation).

The final level of simplification is to remove \emph{all}
unary nodes from the local trees (Fig.~\ref{fig-simplification}D
and E). [And continue]

>>>>

Given a gARG $E$ and the set of samples $S$ we can \emph{resolve} its
ancestry by noting that the resolved interval $Ir$ on an edge $(c,p,I)$
is given by [Rough notes]

\begin{itemize}
\item The ``ancestry interval'' $A_u$ of a node $u$ is the union of
all $R_{u, v}$ for all the children $v$ of $u$. (this isn't quite right)
\item The resolved interval is then the intersection of $A_u$ and the
original inheritance interval $I_v$ on the edge.
\item It would be good of if differentiate between the inheritance intervals
 from the gARG and the ancestry intervals of the resolved gARG.
\end{itemize}

@@

Not all nodes in an ARG are equal. Assuming that we have sequenced the
genomes of the samples, there are intrinsic limits on what can
be known about more ancient nodes, determined by the graph topology
and the flow ancestral material through it.
\emph{Simplification}~\citep{kelleher2018efficient} is the process
of removing nodes and re-writing edges in a gARG to remove
various types of redundancy, while maintaining the crucial properties
that the meaningful topology in the local trees is not affected,
and the relationship between nodes in adjacent trees is maintained.

\begin{figure}
\centering
\vspace{5em}
\includegraphics[width=\linewidth]{illustrations/simplification}
\caption{\label{fig-simplification}
ARG simplification. In all cases the graph visualization is shown on the left
and the equivalent marginal trees on the right.
(A) A gARG simulated from a diploid Wright-Fisher
model. Note that the genome \textsf{K} has three children, and the genomes
\textsf{E} and \textsf{G} simultaneously have multiple parents and multiple children;
neither could be directly represented in a Griffiths-like eARG encoding
(see section XXX). The penultimate local tree is highlighted to aid discussion (see section YYY).
(B) Simplified to remove all
% trying out this terminology - can define in the text
1-connected graph components (e.g., diamonds such as \textsf{JLMN}).
(C) Remove nodes that are unary in all local trees. This removes both recombinant nodes
that have only one child (e.g. \textsf{N}) and nodes that have more
than one child (i.e. are ``common ancestors'') but never represent coalescences
in the local trees (e.g. \textsf{R}).
(D) Rewrite edges to bypass any unary nodes in local trees. Note that this loses lineage
information such as the fact that \textsf{K} is connected to \textsf{A} via \textsf{E}.
}
\end{figure}

% Q: would it help to talk about ``tree vertices'' here or would it
% just be more confusing? I worry that people would think that
% they are two different things then, and it's really important
% that they get that a node-is-a-node

The ideas of ``meaningful'' local tree topologies revolve around the
presence of ``unary'' nodes: a node in the tree that has exactly
one child. Fig.~\ref{fig-simplification} show a series of successive steps of
simplification, starting with a complete gARG simulated under a Wright-Fisher
model, shown both as a graph on the left, and the equivalent set of 6
local trees on the right (Fig.~\ref{fig-simplification}A). Consider the path above node
\textsf{A} in the penultimate tree, highlighted in yellow and covering genomic positions
60 to 70. On this path, before $A$ meets its common ancestor with the other samples at
node \textsf{Q}, it passes through  \textsf{E},  \textsf{J},  \textsf{M},  \textsf{N} and 
\textsf{P}, which in this region of the genome are all unary nodes. This reflects the
passage of this span of genome through the ARG towards the root (see section XXX).
Such unary nodes in a local tree are not usually considered important,
because they provide very little information and are not
detectable by information from the tree. % weak
The nodes in a tree usually mark the points where samples find a
 \emph{common} ancestor, such as node \textsf{Q} in the highlighted
tree. For many purposes, we would therefore consider a path directly
joining sample \textsf{A} to node \textsf{Q} to be of equivalent status to the path
containing the intermediate unary nodes. Simplification is the process of removing
these less informative and (for most computational purposes) inefficient nodes
from the local trees and the corresponding annotated graph.

The first type of simplification that we can perform is to remove
graph topology that is invisible to the samples. The most well
known example of such topology is a so-called
``diamond''~\citep{rasmussen2014genome}
in which the two parent nodes of a recombination immediately
join again into a common ancestor (e.g. \textsf{J}, \textsf{L}, \textsf{M}
and  \textsf{N} in Fig.~\ref{fig-simplification}A).
Unless we are specifically
interested in the recombination event or ancestral genomes,
there is no information in this topology and the diamond can be
replaced by a single edge. More generally, any
subgraph which is singly-connected in both the leafward and
rootward direction (a ``super-diamond'') is non-identifiable and can be
replaced by a single edge. This definition includes the case
of a single node that has one inbound and one outbound edge, such as
nodes \textsf{F} and \textsf{H}.
Fig.~\ref{fig-simplification}B shows the result of this type of
graph topology simplification.

Simplifying away diamonds will remove many unary nodes from the
local trees, but there can still be nodes that are unary in all
of the local trees. In particular, a node can represent a recombinant
with multiple parents in the graph but only a single child (e.g. node \textsf{N}
in Fig.~\ref{fig-simplification}B), or can represent a common ancestor with
multiple children in the graph but in which no local coalescence takes place
(node \textsf{R} in Fig.~\ref{fig-simplification}B).
% point out that this is why Hudson refers to CA nodes, not coalescence nodes
Such nodes are not singly connected in the graph, but are nevertheless unary in
all of the local trees in Fig.~\ref{fig-simplification}B. The operation to remove them,
the results of which are shown in Fig.~\ref{fig-simplification}C,
therefore requires knowledge not just of the graph topology, but also of the
edge annotations. Removal of recombinant nodes can produce graph nodes with
more than 2 parents (e.g. node \textsf{E}); likewise, removal of
common ancestor but non-coalescent nodes can produce graph nodes with
more than 2 children (e.g. node \textsf{S}). These represent a ``stacking up'' of multiple
often indistinguishable events into a single node, something which is impossible
to encode in an eARG.

The remaining nodes are MRCAs of some subset of the samples
at \emph{some} positions along the genome. We still have
some unary nodes in the local trees, but these nodes will
correspond to a coalescence in at least one other
local tree. For example, node  \textsf{K} is unary in the second tree
of Fig.~\ref{fig-simplification}C, but is either binary
or ternary in all subsequent trees (recall this is a Wright-Fisher
simulation). The final level of simplification is to alter the edge annotations
such that, although no nodes are removed from the graph, \emph{all}
unary nodes disappear from the local trees (Fig.~\ref{fig-simplification}D).
It is common for this operation to remove nodes above the top MRCA node
in the tree, resulting in many more root nodes (here, \textsf{Q} and \textsf{K}
become roots). [And continue]
% Note that although this last stage produces much simpler local trees, by
% removing information about the exact paths taken by lineages through
% the graph, it loses potentially important information about shared edges
% between trees. It is not clear whether (c) or (d) leads to a more compact
% representation of the ARG in terms of number of edges.
